---
title: "Function Redundancy Analsysis"
author: "Alexandria igwe"
date: "03/06/2026"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Functional Redundancy
#df.ncbi_eggnog_filtered_ncbisplit_count_merged is a data frame with the counts of taxonomic and functional assignment pairs from all sites
#https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-024-01838-5
#FRIa for within community functional redundancy where high FRIa means many taxa perform a function, low FRIa mean few taxa perform a function, and FRIa = 0 mean 1 taxon performs a function
#FRIb for between community functional redundancy where high FRIb means different taxa perform the same function and low FRIb mean similar taxa perform the same function

```{r}
df.ncbi_eggnog_filtered_ncbisplit_count_merged <- readRDS("~/Desktop/R/CASerpSoilDiversity/input/df.ncbi_eggnog_filtered_ncbisplit_count_merged.RDS")

str(df.ncbi_eggnog_filtered_ncbisplit_count_merged)
names <- c(1:8,10:19)
df.ncbi_eggnog_filtered_ncbisplit_count_merged[,names] <- lapply(df.ncbi_eggnog_filtered_ncbisplit_count_merged[,names] , factor)
str(df.ncbi_eggnog_filtered_ncbisplit_count_merged)
levels(df.ncbi_eggnog_filtered_ncbisplit_count_merged$soil.chem)
dim(df.ncbi_eggnog_filtered_ncbisplit_count_merged)


library(dplyr)
df.ncbi_eggnog_filtered_ncbisplit_count_merged <- dplyr::filter(df.ncbi_eggnog_filtered_ncbisplit_count_merged, soil.chem != ".")
df.ncbi_eggnog_filtered_ncbisplit_count_merged <- droplevels(df.ncbi_eggnog_filtered_ncbisplit_count_merged)
levels(df.ncbi_eggnog_filtered_ncbisplit_count_merged$soil.chem)
dim(df.ncbi_eggnog_filtered_ncbisplit_count_merged)
```

```{r}
library(dplyr)

calculate_FRIa <- function(df, function_col = "eggnog_4", taxon_col = "ncbi_06phy", group_col = "soil.chem", count_col = "n()") {
  df %>%
    filter(!is.na(.data[[taxon_col]])) %>%
    group_by(.data[[group_col]], .data[[function_col]]) %>%
    mutate(Pi = .data[[count_col]] / sum(.data[[count_col]]),
           Pi_lnPi = Pi * log(Pi)) %>%
    summarise(
      FRIa = -sum(Pi_lnPi),
      S = n_distinct(.data[[taxon_col]]),
      FRInorm = FRIa / log(S),
      .groups = "drop"
    )
}

fri_a_results <- calculate_FRIa(df.ncbi_eggnog_filtered_ncbisplit_count_merged)

#write.csv(fri_a_results, "~/Desktop/R/CASerpSoilDiversity/output/fri_a_results_e4n6.csv")

```
```{r}
library(tidyr)
# Prep Pi values per function, taxon, and soil type
fri_b_prep <- df.ncbi_eggnog_filtered_ncbisplit_count_merged %>%
  filter(soil.chem %in% c("serpentine", "nonserpentine"),
         !is.na(ncbi_05)) %>%
  group_by(eggnog_4, soil.chem) %>%
  mutate(Total_n = sum(`n()`),
         Pi = `n()` / Total_n) %>%
  select(eggnog_4, ncbi_06, soil.chem, Pi)

# Pivot to wide: rows = function + taxa; columns = communities
#fri_b_wide <- fri_b_prep %>%
#  pivot_wider(names_from = soil.chem, values_from = Pi, values_fill = list(Pi = 0))
library(data.table)

dt <- as.data.table(fri_b_prep)
fri_b_wide <- dcast(dt, eggnog_4 + ncbi_06 ~ soil.chem, value.var = "Pi", fill = 0, fun.aggregate = sum)

# Calculate FRIb
fri_b <- fri_b_wide %>%
  group_by(eggnog_4) %>%
  summarise(FRIb = 1 - 2 * sum(pmin(serpentine, nonserpentine)) /
                   (sum(serpentine) + sum(nonserpentine)))
write.csv(fri_b, "~/Desktop/R/CASerpSoilDiversity/output/fri_b_e4n6.csv")
```
```{r}
library(dplyr)
library(tidyr)
library(rlang)

calculate_FRIb <- function(df,
                           group_a = "serpentine",
                           group_b = "nonserpentine",
                           function_col = "eggnog_3",
                           taxon_col = "ncbi_05",
                           group_col = "soil.chem",
                           count_col = "n()") {
  
  # Load required library
  library(dplyr)
  library(data.table)
  
  # Rename columns using symbols
  fn_col  <- rlang::sym(function_col)
  tax_col <- rlang::sym(taxon_col)
  grp_col <- rlang::sym(group_col)
  cnt_col <- rlang::sym(count_col)

  # Step 1: Compute Pi
  df_pi <- df %>%
    filter(!!grp_col %in% c(group_a, group_b), !is.na(!!tax_col)) %>%
    group_by(!!fn_col, !!grp_col) %>%
    mutate(Pi = !!cnt_col / sum(!!cnt_col)) %>%
    ungroup() %>%
    select(Function = !!fn_col, Taxa = !!tax_col, Group = !!grp_col, Pi)

  # Step 2: Reshape with data.table
  dt <- as.data.table(df_pi)
  fri_b_wide <- dcast(dt, Function + Taxa ~ Group, value.var = "Pi", fill = 0, fun.aggregate = sum)

  # Step 3: Calculate FRIb
  fri_b_result <- fri_b_wide[, .(
    FRIb = 1 - 2 * sum(pmin(get(group_a), get(group_b))) /
              (sum(get(group_a)) + sum(get(group_b)))
  ), by = Function]

  return(fri_b_result)
}

fri_b_results <- calculate_FRIb(df.ncbi_eggnog_filtered_ncbisplit_count_merged)
```
```{r}
df_duplicates <- dt %>%
  group_by(eggnog_3, ncbi_06, soil.chem) %>%
  filter(n() > 1) %>%
  arrange(eggnog_3, ncbi_06, soil.chem)
```
```{r}
fri_a_results_e3n5$soil.chem <- as.factor(fri_a_results_e3n5$soil.chem)
fri_a_results_e3n5$eggnog_3 <- as.factor(fri_a_results_e3n5$eggnog_3)
res_aov_soil.chem <- aov(FRInorm ~ soil.chem, data = fri_a_results_e3n5)
summary(res_aov_soil.chem)

res_aov_e3 <- aov(FRInorm ~ eggnog_3, data = fri_a_results_e3n5)
summary(res_aov_e3)
TukeyHSD(res_aov_e3)

res_aov_soil.chem_e3 <- aov(FRInorm ~ soil.chem*eggnog_3, data = fri_a_results_e3n5)
summary(res_aov_soil.chem_e3)
tukey_res_aov_soil.chem_e3<-TukeyHSD(res_aov_soil.chem_e3)
tukey_res_aov_soil.chem_e3 <- as.data.frame(tukey_res_aov_soil.chem_e3)
tukey_res_aov_soil.chem_e3$`soil.chem:eggnog_3`
write.csv(tukey_res_aov_soil.chem_e3$`soil.chem:eggnog_3`,"~/Desktop/R/CASerpSoilDiversity/output/Tukey_res_aov_soil.chem_e3.csv")


#                   Df Sum Sq Mean Sq
#soil.chem           1  0.014 0.01363
#eggnog_3           21  3.915 0.18644
#soil.chem:eggnog_3 21  0.032 0.00154
```
```{r}
library(lme4)
library(lmerTest)  # for p-values

res_lmer <- lmer(FRInorm ~ soil.chem + (1|eggnog_3), data = fri_a_results_e3n5)
summary(res_lmer)
```
```{r}
library(dplyr)

res_per_func <- fri_a_results_e3n5 %>%
  group_by(eggnog_3) %>%
  summarise(
    test = list(
      tryCatch(
        t.test(FRInorm ~ soil.chem, data = cur_data()),
        error = function(e) NULL
      )
    ),
    .groups = "drop"
  ) %>%
  mutate(
    p.value = sapply(test, function(x) if (!is.null(x)) x$p.value else NA),
    p.adj = p.adjust(p.value, method = "fdr")
  ) %>%
  select(eggnog_3, p.value, p.adj)

res_per_func
```
