---
title: "Parsing DIAMOND+MEGAN ncbi table"
author: "Alexandria igwe"
date: "12/25/2022"
output: html_document
---
#The -r2c or --read2class option in MEGAN (Metagenome Analyzer) is used to export read-to-class assignments (taxonomic or functional) from a MEGAN RMA6 file
#This flag was used to export ncbi and eggNOG functional assignments
#The assignments for each site were combined into one large data frame. A subset of this data frame was used to test the workflow, then the workflow was applied to the entire data frame.
#There may be cleaner ways to do this, but essentially a unique list of ncbi assignments was obtained then used to create a taxonomy table for phyloseq.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#subset all data.frame for testing
```{r}
#daa2info_r2c_ncbi_eggnog_df_all is the combined output form the --read2class flag
ncbi_eggnog_df.5000 <- daa2info_r2c_ncbi_eggnog_df_all[sample(nrow(daa2info_r2c_ncbi_eggnog_df_all), 5000), ]
class(ncbi_eggnog_df.5000)

proportion <- 0.4  # 10% of each group
ncbi_eggnog_df.10 <- daa2info_r2c_ncbi_eggnog_df_all_assigned %>%
    group_by(daa2info_r2c_ncbi_eggnog_df_all_assigned$sample_name) %>%
    sample_frac(proportion)

ncbi_eggnog_df.5000 <- ncbi_eggnog_df.10
ncbi_eggnog_df.5000 <- daa2info_r2c_ncbi_eggnog_df_all_assigned

ncbi_eggnog_df.5000_assigned <- subset(ncbi_eggnog_df.5000, eggnog != "eggNOG;Not assigned;")
ncbi_eggnog_df.5000_assigned <- subset(ncbi_eggnog_df.5000_assigned, ncbi != "NCBI; Not assigned;")

ncbi_eggnog_df.5000_assigned <- ncbi_eggnog_df.5000

saveRDS(ncbi_eggnog_df.10,"~/CASerpSoil/input/ncbi_eggnog_df_10.RDS")
```
#create ncbi only phyloseq object
```{r}
#split ncbi assignment into groups for phyloseq tax table
unique_ncbi_eggnog_df.5000 <- as.data.frame(unique(ncbi_eggnog_df.5000_assigned$ncbi))
colnames(unique_ncbi_eggnog_df.5000)[1] <- "ncbi"
unique_split_ncbi_eggnog_df.5000 <- cSplit(unique_ncbi_eggnog_df.5000, "ncbi", sep = ";", direction = "wide", fixed = FALSE)
unique_split_ncbi_eggnog_df.5000 <- as.data.frame(unique_split_ncbi_eggnog_df.5000)
rownames(unique_split_ncbi_eggnog_df.5000) <- c(unique_ncbi_eggnog_df.5000$ncbi)
```
```{r}
#split_ncbi_eggnog_df.5000 <- cSplit(ncbi_eggnog_df.5000_assigned, "ncbi", sep = ";", direction = "wide", fixed = FALSE)

# Initialize a list to store the unique value data frames
#unique_lists <- list()

# Loop through the column names from ncbi_01 to ncbi_16
#for (i in 1:15) {
  # Construct the column name
#  column_name <- paste0("ncbi_", sprintf("%02d", i))
  
  # Check if the column exists in the data frame
#  if (column_name %in% names(split_ncbi_eggnog_df.5000)) {
    # Extract unique values and store in a data frame
#    unique_values <- as.data.frame(unique(split_ncbi_eggnog_df.5000[[column_name]]))
    
    # Assign the unique values data frame to the list with a specific name
#    unique_lists[[column_name]] <- unique_values
#  } else {
#   warning(paste("Column", column_name, "does not exist in the data frame."))
#  }
#}

# Print the unique lists
#unique_lists
```
```{r}
#subset bacteria
split_ncbi_eggnog_df.5000.bacteria <- subset(unique_split_ncbi_eggnog_df.5000, ncbi_03 == "Bacteria")
#split_ncbi_eggnog_df.5000.bacteria <- subset(split_ncbi_eggnog_df.5000, ncbi_03 == "Bacteria")

#same as below, except columns are shifted
colnames(split_ncbi_eggnog_df.5000.bacteria)[3] <- "Domain"
colnames(split_ncbi_eggnog_df.5000.bacteria)[4] <- "Phylum"
colnames(split_ncbi_eggnog_df.5000.bacteria)[5] <- "Class"
colnames(split_ncbi_eggnog_df.5000.bacteria)[6] <- "Order"
colnames(split_ncbi_eggnog_df.5000.bacteria)[7] <- "Family"
colnames(split_ncbi_eggnog_df.5000.bacteria)[8] <- "Genus"
colnames(split_ncbi_eggnog_df.5000.bacteria)[9] <- "Species"

#change column name to "Domain"
#colnames(split_ncbi_eggnog_df.5000.bacteria)[7] <- "Domain"
#colnames(split_ncbi_eggnog_df.5000.bacteria)[8] <- "Phylum"
#colnames(split_ncbi_eggnog_df.5000.bacteria)[9] <- "Class"
#colnames(split_ncbi_eggnog_df.5000.bacteria)[10] <- "Order"
#colnames(split_ncbi_eggnog_df.5000.bacteria)[11] <- "Family"
#colnames(split_ncbi_eggnog_df.5000.bacteria)[12] <- "Genus"
#colnames(split_ncbi_eggnog_df.5000.bacteria)[13] <- "Species"

#create a bacteria only data frame and adjust taxonomy names
a.split_ncbi_eggnog_df.5000.bacteria <- subset(split_ncbi_eggnog_df.5000.bacteria, !grepl("group", split_ncbi_eggnog_df.5000.bacteria$Phylum))
b.split_ncbi_eggnog_df.5000.bacteria <- subset(split_ncbi_eggnog_df.5000.bacteria, grepl("group", split_ncbi_eggnog_df.5000.bacteria$Phylum))
colnames(b.split_ncbi_eggnog_df.5000.bacteria)[4] <- "Superphylum"
colnames(b.split_ncbi_eggnog_df.5000.bacteria)[5] <- "Phylum"
colnames(b.split_ncbi_eggnog_df.5000.bacteria)[6] <- "Class"
colnames(b.split_ncbi_eggnog_df.5000.bacteria)[7] <- "Order"
colnames(b.split_ncbi_eggnog_df.5000.bacteria)[8] <- "Family"
colnames(b.split_ncbi_eggnog_df.5000.bacteria)[9] <- "Genus"
colnames(b.split_ncbi_eggnog_df.5000.bacteria)[10] <- "Species"
b.split_ncbi_eggnog_df.5000.bacteria[, c(11)] <- NULL

#same as above, but columns shifted
#colnames(b.split_ncbi_eggnog_df.5000.bacteria)[8] <- "Superphylum"
#colnames(b.split_ncbi_eggnog_df.5000.bacteria)[9] <- "Phylum"
#colnames(b.split_ncbi_eggnog_df.5000.bacteria)[10] <- "Class"
#colnames(b.split_ncbi_eggnog_df.5000.bacteria)[11] <- "Order"
#colnames(b.split_ncbi_eggnog_df.5000.bacteria)[12] <- "Family"
#colnames(b.split_ncbi_eggnog_df.5000.bacteria)[13] <- "Genus"
#colnames(b.split_ncbi_eggnog_df.5000.bacteria)[14] <- "Species"
#b.split_ncbi_eggnog_df.5000.bacteria[, c(15)] <- NULL

split_ncbi_eggnog_df.5000.bacteria2 <- merge(a.split_ncbi_eggnog_df.5000.bacteria, b.split_ncbi_eggnog_df.5000.bacteria, by = c("ncbi_01","ncbi_02","Domain", "Phylum", "Class", "Order","Family", "Genus", "Species"), all = TRUE)
split_ncbi_eggnog_df.5000.bacteria2[, c(10:11)] <- NULL

#unique_group <- as.data.frame(unique(split_ncbi_eggnog_df.5000.bacteria2$Class))

# Specify new column order
#"seq_id","eggnog","sample_name","sample_type",
new_order <- c("ncbi_01", "ncbi_02", "Domain", "Superphylum","Phylum", "Class", "Order", "Family", "Genus", "Species")
# Reorder columns
split_ncbi_eggnog_df.5000.bacteria2 <- split_ncbi_eggnog_df.5000.bacteria2[, new_order]

# Use apply and paste to combine the columns
split_ncbi_eggnog_df.5000.bacteria2$ncbi <- apply(split_ncbi_eggnog_df.5000.bacteria2[, c(new_order)], 1, paste, collapse = "; ")

# Define custom function to paste values, ignoring NA
paste_ignore_na <- function(x) {
  # Filter out NA values
  x <- na.omit(x)
  # Paste values together, separated by semicolon
  #paste(x, collapse = "; ")
  # Paste values together, separated by a semicolon, and add a trailing semicolon
  result <- paste(x, collapse = "; ")
  if (nchar(result) > 0) {  # Only add a semicolon if result is not empty
    result <- paste0(result, ";")
  }
  return(result)
}

# Combine specified columns into one, separated by a semicolon, ignoring NAs
split_ncbi_eggnog_df.5000.bacteria2$ncbi <- apply(
  split_ncbi_eggnog_df.5000.bacteria2[, new_order],
  1,
  paste_ignore_na
)

split_ncbi_eggnog_df.5000.bacteria2 <- subset(split_ncbi_eggnog_df.5000.bacteria2, !grepl("unclassified Bacteria", split_ncbi_eggnog_df.5000.bacteria2$Phylum))

split_ncbi_eggnog_df.5000.bacteria2 <- subset(split_ncbi_eggnog_df.5000.bacteria2, !grepl("environmental samples <bacteria,superkingdom Bacteria>", split_ncbi_eggnog_df.5000.bacteria2$Phylum))

daa2info_ps_tax <- split_ncbi_eggnog_df.5000.bacteria2

# Remove duplicate rows based on the 'ncbi' column
daa2info_ps_tax <- split_ncbi_eggnog_df.5000.bacteria2 %>%
  distinct(ncbi, .keep_all = TRUE)

# Now, set the row names to the unique values of the 'ncbi' column
rownames(daa2info_ps_tax) <- daa2info_ps_tax$ncbi

# Optionally, remove the 'ncbi' column if it is no longer needed
# daa2info_ps_tax$ncbi <- NULL

# View the resulting data frame with unique row names
head(daa2info_ps_tax)

class(daa2info_ps_tax)
```
##OTU table
```{r}
#for otu table
#https://datacarpentry.org/R-genomics/04-dplyr.html
#https://stackoverflow.com/questions/9617348/reshape-three-column-data-frame-to-matrix-long-to-wide-format

#aggregate information
df.ncbi_count <- ncbi_eggnog_df.5000_assigned %>%
  group_by(ncbi,sample_name) %>%
  summarize(n())
unique(df.ncbi_count$sample_name)

daa2info_ps_otu_tmp <- acast(df.ncbi_count, df.ncbi_count$ncbi~df.ncbi_count$sample_name, value.var="n()")
daa2info_ps_otu_tmp[is.na(daa2info_ps_otu_tmp)] <- 0
View(daa2info_ps_otu_tmp[1:4,])
daa2info_ps_otu_tmp <- as.data.frame(daa2info_ps_otu_tmp)
colnames(daa2info_ps_otu_tmp)
rownames(daa2info_ps_otu_tmp)
#daa2info_ps_otu_tmp$ncbi <- rownames(daa2info_ps_otu_tmp)

#https://github.com/joey711/phyloseq/issues/1020
OTU.ncbi = otu_table(daa2info_ps_otu_tmp, taxa_are_rows = TRUE)
TAX.ncbi = tax_table(as.matrix(daa2info_ps_tax))
META.ncbi = sample_data(daa2info_ps_meta)
physeq.ncbi = phyloseq(OTU.ncbi, TAX.ncbi, META.ncbi)
physeq.ncbi
##Change taxa names to 'Seq#'
taxa_names(physeq.ncbi) <- paste0("Seq", seq(ntaxa(physeq.ncbi)))
physeq.ncbi
```
#decontam - ncbi
```{r}
#decontam
sample_data(physeq.ncbi)$is.neg <- sample_data(physeq.ncbi)$soil.chem == "."
contamdf.prev.ncbi <- isContaminant(physeq.ncbi, method="prevalence", neg="is.neg")
table(contamdf.prev.ncbi$contaminant)
head(which(contamdf.prev.ncbi$contaminant))
contamdf.prev.ncbi05 <- isContaminant(physeq.ncbi, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev.ncbi05$contaminant)
# Make phyloseq object of presence-absence in negative controls and true samples
ps.pa.ncbi <- transform_sample_counts(physeq.ncbi, function(abund) 1*(abund>0))
ps.pa.ncbi.neg <- prune_samples(sample_data(physeq.ncbi)$soil.chem == ".", ps.pa.ncbi)
ps.pa.ncbi.pos <- prune_samples(sample_data(physeq.ncbi)$soil.chem != ".", ps.pa.ncbi)
# Make data.frame of prevalence in positive and negative samples
df.pa.ncbi <- data.frame(pa.ncbi.pos=taxa_sums(ps.pa.ncbi.pos), pa.ncbi.neg=taxa_sums(ps.pa.ncbi.neg),
                      contaminant=contamdf.prev.ncbi$contaminant)
ggplot(data=df.pa.ncbi, aes(x=pa.ncbi.neg, y=pa.ncbi.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
physeq.ncbi.decontam <- prune_taxa(!contamdf.prev.ncbi05$contaminant, physeq.ncbi)
physeq.ncbi.decontam
```
